---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---

## Loading and preprocessing the data
```{r}
library(knitr)
library(plyr)
library(dplyr)
library(lubridate)
library(Hmisc)
unzip("activity.zip")
## Read in data
activityDataRaw <- read.csv("activity.csv")
activityDataTidy <- activityDataRaw
##Normalize interval variable to a continous interval variable in minutes
activityDataTidy$interval <- (activityDataRaw$interval %/% 100 * 60) + (activityDataRaw$interval %% 100)
```

## What was the mean total number of steps taken per day?
```{r}
dailySteps <- activityDataTidy %>%
  group_by(date) %>%
  summarise(sum(steps, na.rm=TRUE))

hist(dailySteps$sum, main = paste("Histogram of Total Number of Steps Taken Each Day"), xlab="Total Number of Steps per Day", breaks=10)
meanSteps <- mean(dailySteps$sum, na.rm = TRUE)
medianSteps <- median(dailySteps$sum, na.rm = TRUE)
```

**Mean Daily Steps**

The **mean number of daily steps** is **`r meanSteps`.**

**Median Daily Steps**

The **median number of daily steps** is **`r medianSteps`.**

## What is the average daily activity pattern?
```{r}
meanStepsPerInterval <- activityDataTidy %>%
  group_by(interval) %>%
  summarise(mean(steps, na.rm = TRUE))

plot(meanStepsPerInterval$interval, meanStepsPerInterval$mean, type="l", ylab="Mean Number of Steps", xlab="Time of day [min] - 5 min intervals", main="Mean Number of Steps per 5 Minute Interval throughout the Day")

maxIndex <- which.max(meanStepsPerInterval$mean)
maxInterval <- meanStepsPerInterval$interval[maxIndex] 
```

The 5-minute interval, on average across all the days in the dataset, containing the maximum number of steps is: `r maxInterval` - `r  maxInterval + 5` min of the day

## Inputing missing values
```{r}
missingObservations <- sum(is.na(activityDataTidy$steps))
```

In total there were `r missingObservations` missing values in the original dataset.  

These observations were substituted with the mean number of steps per interval, as calculated ignoring the missing values.

This substitution produces an adjusted set.  
```{r}
activityDataTidy2 <- activityDataTidy
activityDataTidy2$steps <- impute(activityDataTidy2$steps,mean)

dailySteps2 <- activityDataTidy2 %>%
  group_by(date) %>%
  summarise(sum(steps, na.rm=TRUE))

hist(dailySteps2$sum, main = paste("Histogram of Total Number of Steps Taken Each Day"), xlab="Total Number of Steps per Day", breaks=10)
meanSteps2 <- mean(dailySteps2$sum, na.rm = TRUE)
medianSteps2 <- median(dailySteps2$sum, na.rm = TRUE)
```

In consideration of the new data set, the following data is returned.

**Mean Daily Steps**

The **mean number of daily steps** is **`r meanSteps2`.**

**Median Daily Steps**

The **median number of daily steps** is **`r medianSteps2`.**

Making adjustments for the missing values has dropped both the median and mean for the set.

## Are there differences in activity patterns between weekdays and weekends?
```{r}
 activityDataTidy2$dayofweek <- as.factor(weekdays(ymd(activityDataTidy2$date)))
 activityDataTidy2$dayofweek <- revalue(activityDataTidy2$dayofweek, c("Monday"="weekday","Tuesday"="weekday","Wednesday"="weekday","Thursday"="weekday","Friday"="weekday", "Saturday"="weekend", "Sunday"="weekend"))

dailySteps3 <- activityDataTidy2 %>%
  group_by(dayofweek, interval) %>%
  summarise(mean(steps)) 

  xyplot(dailySteps3$mean ~ interval | dayofweek, dailySteps3, 
         type="l",
         ylab="Mean Number of Steps", 
         xlab="Time of day [min] - 5 min intervals", 
         main="Average Number of Steps per 5 Minute Interval throughout the Day",
         layout=c(1,2))
```
